The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-31 paper/paper.tex
paper: prose improvements near start

Introduced by Reynolds in the context of programming language semantics~\citep{reynolds83}, relational parametricity happens to be connected to several similar concepts in logic and mathematics. Associating binary propositional relations to types in its original formulation, it can also associate proof-relevant relations~\cite{GhaniNordvallOrsanigo16} or unary relations, that is predicates (proof-irrelevant case) or families (relevant case). These unary or binary, relevant or proof-irrelevant relations express what it means to belong in a type (unary case) or to be equal in a type (binary case) from an observational point of view. For instance, the canonical predicate associated to $A \rightarrow B$ is the predicate over functions $f$ in $A \rightarrow B$ such that, for all $a$ in the canonical predicate associated to $A$, $f(a)$ is in the canonical predicate associated to $B$. And, similarly, the canonical relation associated to $A \rightarrow B$ is the relations over functions $f$ and $f'$ in $A \rightarrow B$ such that, for all $a$ and $a'$ related through the canonical relation associated to $A$, $f(a)$and $f'(a')$ are related through the canonical relation associated to $B$.

From the point of view of realisability, unary parametricity over a language is a realisability using the language itself as language of realisers. Like parametricity, realisability can be expressed in a \emph{syntactic} way, that is as an interpretation from one language to another (i.e. mapping an expression $t$ of $\vdash A$ in some formal system to a derivation $t^*$ of $\vdash' |t| \textsf{r} A$ in some other formal system, for some program $|t|$ extracted from $t$ -- namely $t$ itself in the case of parametricity), or a \emph{semantic} way\footnote{This distinction between syntactic realisability and semantic realisability is itself connected to the difference between relative and absolute consistency. In the relative case, consistency of a formal system is reduced to the consistency of another formal system, and this can be done in a very weak logical system such as Primitive Recursive Arithmetic (PRA). On the other side, in the absolute case, consistency of a formal system is reduced to the consistency of the language of discourse, which, by G\"odel incompleteness theorem, if actually consistent, necessarily has to be as strong as the modelled formal system.}, that is as an interpretation in the language of discourse, which we could call in term the ``metalanguage'' (i.e. interpreting $A$ as a set $\textbf{A}$ or a relation $=_A$ of the language of discourse and interpreting a derivation $t$ of $\vdash A$ as an expression $|t|$ such that, respectively, $|t| \in \textbf{A}$ or $|t| =_A |t|$). Thus, the comparison between parametricity and realisability includes the case of \emph{reducibility}~\cite{Tait}, which is a semantic form of realisability, also known as \emph{glueing} in the context of category theory. In the same vein, binary parametricity over a language can be seen as defining \emph{logical relations}~\cite{Plotkin} over itself, since logical relations are a binary form of reducibility.

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form. To express it, let us use a type-theoretic language featuring a type of propositions $\Prop$, a type of sets $\U$, as well, as the standard $\Sigma$ and $\Pi$ constructions. In such language, predicates have type $A \rightarrow \Prop$, families have type $A \rightarrow \U$, relations have type $A \times A \rightarrow \Prop$, relevant relations -- sometimes called \emph{correspondences} -- have type $A \times A \rightarrow \U$. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\U. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \U$ and $\Sigma B:\U. (B \rightarrow A)$ (proof-relevant case), for $A:\U$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

>>>>

Introduced by Reynolds in the context of programming language semantics~\citep{reynolds83}, relational parametricity happens to be connected to several similar concepts in logic and mathematics. Associating binary propositional relations to types in its original formulation, it can also associate proof-relevant relations~\citep{GhaniNordvallOrsanigo16} or unary relations, that is predicates (proof-irrelevant case) or families (relevant case). These unary or binary, relevant or proof-irrelevant relations express what it means to belong in a type (unary case) or to be equal in a type (binary case) from an observational point of view. For instance, the canonical predicate associated to $A \rightarrow B$ is the predicate over functions $f$ in $A \rightarrow B$ such that, for all $a$ in the canonical predicate associated to $A$, $f(a)$ is in the canonical predicate associated to $B$. And, similarly, the canonical relation associated to $A \rightarrow B$ is the relations over functions $f$ and $f'$ in $A \rightarrow B$ such that, for all $a$ and $a'$ related through the canonical relation associated to $A$, $f(a)$and $f'(a')$ are related through the canonical relation associated to $B$.

From the point of view of realisability, unary parametricity over a language is a realisability using the language itself as language of realisers. Like parametricity, realisability can be expressed in a \emph{syntactic} way, that is as an interpretation from one language to another (i.e. mapping an expression $t$ of $\vdash A$ in some formal system to a derivation $t^*$ of $\vdash' |t| \textsf{r} A$ in some other formal system, for some program $|t|$ extracted from $t$ -- namely $t$ itself in the case of parametricity), or a \emph{semantic} way\footnote{This distinction between syntactic realisability and semantic realisability is itself connected to the difference between relative and absolute consistency. In the relative case, consistency of a formal system is reduced to the consistency of another formal system, and this can be done in a very weak logical system such as Primitive Recursive Arithmetic (PRA). On the other side, in the absolute case, consistency of a formal system is reduced to the consistency of the language of discourse, which, by G\"odel incompleteness theorem, if actually consistent, necessarily has to be as strong as the modelled formal system.}, that is as an interpretation in the language of discourse, which we could call in term the ``metalanguage'' (i.e. interpreting $A$ as a set $\textbf{A}$ or a relation $=_A$ of the language of discourse and interpreting a derivation $t$ of $\vdash A$ as an expression $|t|$ such that, respectively, $|t| \in \textbf{A}$ or $|t| =_A |t|$). Thus, the comparison between parametricity and realisability includes the case of \emph{reducibility}~\citep{Tait67}, which is a semantic form of realisability, also known as \emph{glueing} in the context of category theory. In the same vein, binary parametricity over a language can be seen as defining \emph{logical relations}~\citep{Plotkin73} over itself, since logical relations are a binary form of reducibility.

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form. To express it, let us use a type-theoretic language featuring a type of propositions $\Prop$, a type of sets $\U$, as well, as the standard $\Sigma$ and $\Pi$ constructions. In such language, predicates have type $A \rightarrow \Prop$, families have type $A \rightarrow \U$, relations have type $A \times A \rightarrow \Prop$, relevant relations -- sometimes called \emph{correspondences} -- have type $A \times A \rightarrow \U$. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\U. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \U$ and $\Sigma B:\U. (B \rightarrow A)$ (proof-relevant case), for $A:\U$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in~\cite{atkey14}, who also suggested to iterate parametricity.
